Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-10 draft.tex
first draft with 8 pages

In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.

@@

We incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.

@@

The extracted lumbar spine shape was correlated with subject size.

@@

	\item 113 categorical variables divided into
		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status

@@

All categorical variables are converted into binary \emph{dummy variables}, indicate the presence or absence of a categorical variable manifestation. %with multiple manifestations into several dichotomous variables.

@@

For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.

@@

This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.

@@

%% % TODO: [high] Add Link
We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at TODO:LINK).

@@

The \texttt{GPLOM} shows very similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.

@@

The described visualization provides a good overview of the classification results.

@@

We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.

@@

This is realized by clicking on an entry in the visualization, which then interactively displays the corresponding decision tree in detail.

@@

The result however, is probably overfitted to the data due to tree sizes between $14$ and $16$.

@@

Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$, but only using large decision trees with over 20 rules.

@@

The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.

@@

The resulting tree consists of $8$ rules and has an error of $0.35$.

>>>>

In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP} subjects \cite{Klemm2013VMV}.

@@

We incorporate hierarchical agglomerative clustering to derive shape groups, yielding average shape groups and outliers.

@@

%The extracted lumbar spine shape was correlated with subject size.

@@

	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
% 		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
% 		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status

@@

All categorical variables are converted into binary \emph{dummy variables}, indicating the presence or absence of a categorical variable manifestation. %with multiple manifestations into several dichotomous variables.

@@

For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables, allowing to determine which manifestation can be described best using the image-based variables

@@

%This allows to determine which manifestation of a variable can be described best using the image data variables.
%This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.

@@

@@

We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at \url{ivapp15.dnsalias.com}).
The \texttt{GPLOM} shows similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.

@@

The visualization provides a good overview of the classification results.

@@

%We still want to be able to display \emph{details-on-demand} and examine a decision tree in detail.

@@

\emph{Details-on-demand} are displayed by clicking on an entry in the visualization, which then displays the corresponding decision tree in detail.

@@

The result is overfitted to the data due to tree sizes between $14$ and $16$.

@@

Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$ using large decision trees with over 20 rules.

@@

Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same features with a tree consisting of $8$ rules and has an error of $0.35$
%The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.

@@

%The resulting tree consists of $8$ rules and has an error of $0.35$.
